\Large\textbf{PixelBytes: Catching Unified Representation for Multimodal Generation}

### 2024-09-13 docs/paper/article.tex
run to publish

\begin{abstract}
This report introduces PixelBytes, a novel approach for unified multimodal representation learning. Our method captures diverse inputs in a single, cohesive representation, enabling emergent properties for multimodal sequence generation, particularly for text and pixelated images. Inspired by state-of-the-art sequence models such as Image Transformers, PixelCNN, and Mamba-Bytes, PixelBytes aims to address the challenges of integrating different data types. We explore various model architectures, including Recurrent Neural Networks (RNNs), State Space Models (SSMs), and Attention-based models, focusing on bidirectional processing and our innovative PxBy embedding technique. Our experiments, conducted on a specialized PixelBytes Pokémon dataset, demonstrate that bidirectional sequence models with PxBy embedding and convolutional layers can generate coherent multimodal sequences. This work contributes to the advancement of integrated AI models capable of understanding and generating multimodal data in a unified manner. Code is available at \url{https://github.com/fabienfrfr/PixelBytes}.
\end{abstract}

\let\thefootnote\relax\footnotetext{$^{*}$ Corresponding author : Fabien Furfaro (\url{fabien.furfaro@gmail.com}).}

\begin{figure}[h]
\centering
\includegraphics[width=0.8\linewidth]{approach_generation.png}
\caption{PixelBytes Pokémon generation process. Left: Image and text tokenization with embedding window for modality unification. Center: Sequence generation using PxBy Embedding. Right: Generated pixelated image and text description. Red box shows a possible generation example. Note: The model currently has some inaccuracies in image generation and may invent words, indicating areas for future improvement.}
\label{fig:approach_generation}
\end{figure}

Amélioration de l'idée : il y a trop de sequence qui se repete ! finaleemnt, la proposition de l'embedding n'est pas bonne ! les resultats avec plus d'approfondissement ne montre pas vraiment que l'embedding avec un convolution est efficace. plutot utiliser les 6 positions specifique (dim entree = dim sortie) et agrandire l'embedding, la génération est mieux (voir Figure 1) ! de meme simplification du code necessaire ! le tokenizer doit etre plus flexible, si on a besoin de générer les donner en plus, ca devient compliqué ! il faut que le tokenizer fasse tout ! (ou utiliser l'egalité entre la dimension de sortie et de l'entrée pour faire un vrai modele autoregressif) Donc dans l'article, il faut recommancer, il y aura donc deux partie ! d'abords la premiere version, pour exploration --> on voit que l'embedding n'est pas suffisant, donc on passe à la nouvelle approche ! (ne pas oublier de comparer avec ImageBind, l'idée n'est pas d'avoir des intermediaire qui ne permette pas l'emergence de propriété)

L'idée n'est pas de faire un modele "ultra performant", sinon pour ces taches, les modeles de diffsion sont mieux (Why Does Diffusion Work Better than Auto-Regression?) maais plutot des mdoele generaux de "fondation" qui ont appris toutes les modalités d'un coup. Ici on ne test que les modele auto-regressif, mais des approche comme Diffusion-LM pourrait etre une autre proposition (et compatible avec notre tokenizer)

@@

Recent advancements in artificial intelligence have led to increasingly generalist models, not by combining multiple specialized components, but by giving simple tasks to models where emergent properties—complex behaviors that arise from simpler underlying rules—appear. This is the case with generative language models like GPT \cite{brown2020language}. However, these models are limited by their focus on language alone, failing to capture the full complexity of multimodal understanding \cite{huang2023language}. To address this limitation, researchers have explored combining LLMs with other modalities \cite{liu2024visual}. But this brings us back to the initial problem, as it often results in specialized model combinations without allowing for new emergent properties. We propose "PixelBytes Embedding", a novel approach enabling unified training across modalities by capturing diverse inputs in a single, cohesive representation.

Multimodal sequence generation, which involves the creation of coherent outputs combining various data types such as text, images, and numerical sequences, presents a significant challenge in artificial intelligence \cite{baltruvsaitis2018multimodal}. While models like GPT have excelled in text generation \cite{brown2020language}, there's a growing need for unified approaches that can seamlessly handle varied data types. Building upon these findings, PixelBytes aims to address the challenge of unified text and image generation by proposing a model capable of producing mixed sequences of text and images in a coherent and unified manner. We draw inspiration from state-of-the-art sequence models, including Image Transformers \cite{parmar2018image}, PixelCNN \cite{van2016conditional}, and recent developments in bytes generation by mamba architectures \cite{wang2024mambabyte}.

Our research explores various architectures including Recurrent Neural Networks (RNNs), State Space Models (SSMs) \cite{dao2024transformers}, and Attention-based models, focusing on the effectiveness of bidirectional processing \cite{zhu2024vision}, novel embedding techniques (particularly PxBy embedding, which unifies pixel and byte-level representations), the impact of convolutional layers, effects of model depth, input dimensionality, and size. Our experiments reveal that bidirectional sequence models with PxBy embedding and convolutional layers demonstrate good performance in generating coherent sequences.

Our results show that it is possible to train sequence models with unified embeddings for both text and pixel completion. This paper presents our methodology to construct unified sequence data from text and image, experimental results, and analysis, contributing to the development of our sequence generation embedding models. The proposed PixelBytes Embedding approach may offer valuable insights for the development of multimodal AI, potentially leading to models that can better understand and generate various types of data.

\section{Model Architecture}

\subsection{Overview}
The PixelBytes architecture is designed to seamlessly integrate multimodal data for unified sequence generation. At its core, the model incorporates two key innovative components: a specialized tokenizer sequence constructor and a unified multimodal embedding technique called PxByEmbed. The tokenizer sequence constructor is engineered to process both pixelated images and text at a byte level, enabling a consistent representation across modalities. This is complemented by PxByEmbed, our novel embedding approach that creates a unified representation for both pixel and byte data in a single, coherent space. This architecture draws inspiration from recent advancements in multimodal learning \cite{baltruvsaitis2018multimodal} and efficient sequence modeling \cite{vaswani2017attention}, allowing PixelBytes to capture intrinsic relationships between visual and textual information effectively.

\subsection{Dataset Construction}

Image captioning datasets, while combining visual and textual modalities, prove unsuitable for joint text and image generation due to limited text content and difficulties in interpreting pixelated versions of high-resolution images. To address these limitations, we developed a specialized Pokémon dataset, offering advantages such as a long-standing franchise history, pixelated designs, rich descriptive text, and over 1000 unique Pokémon.

We constructed our dataset by web scraping Pokémon miniatures and descriptions from Pokepedia using Beautiful Soup \cite{richardson2007beautiful}, maintaining a 2/3 text to 1/3 image ratio. For image quantization, we employed a 55-color palette inspired by the NES, creating tokens representing different color combinations. This approach translates visual information into a format suitable for sequence modeling. The quantization and pixelization process used OpenCV and scikit-image \cite{bradski2000opencv, van2014scikit}. To ensure balanced representation, we adjusted the number of image and text tokens to 113 indices for all entries, allowing the model to learn equally from both visual and textual information. The final dataset, balancing text and pixelated images, is available on the Hugging Face Datasets Hub for reproducibility at \url{https://huggingface.co/datasets/ffurfaro/PixelBytes-Pokemon}.

\section{Multimodal Embedding Algorithm}

\subsection{PxByEmbed: Multimodal Embedding Algorithm}

At the core of our approach is the PxByEmbed algorithm, which represents mixed sequences of text and pixelated images in a unified manner. This algorithm builds upon existing embedding techniques by incorporating spatial adaptivity, allowing for representation of both textual and visual information. PxByEmbed is designed to address the specific needs of pixel-level image representation and byte-level text encoding. The algorithm uses a learned embedding matrix to map each token (text or image) to a vector space, while maintaining spatial relationships for image tokens. This approach allows our model to handle both modalities within a single framework.

\begin{algorithm}[ht]
\caption{PxByEmbed: Multimodal Embedding Algorithm (k=3)
\newline
\textbf{Input:} $V$: vocabulary size, $D$: embedding dimension
\newline
\textbf{Output:} Embedded representation $\mathbf{E} \in \mathbb{R}^{B \times L \times D}$
\newline
\textbf{Note:} $\mathbf{X}_{emb} \in \mathbb{R}^{B \cdot L \times E_{int} \times k \times k}$, 
$\mathbf{X}_{flat} \in \mathbb{R}^{B \cdot L \times E_{int}k^2}$, 
$\mathbf{X}_{proj} \in \mathbb{R}^{B \cdot L \times D}$
}
\begin{algorithmic}[0]
\State \textbf{Initialize:}
\State $k \gets 3$
\State $E_{int} \gets \max(9, \lfloor D / k^2 \rfloor)$
\State $\mathbf{\alpha} \in \mathbb{R}^{1 \times 1 \times k \times k}$ 
\State $\mathbf{W}_{emb} \in \mathbb{R}^{V \times E_{int}}$ 
\State $\mathbf{W}_{proj} \in \mathbb{R}^{E_{int}k^2 \times D}$ 
\State $\mathbf{W}_{patch} \in \mathbb{R}^{E_{int} \times E_{int} \times k \times k}$

@@

\subsection{Managing Transitions}

PixelBytes uses a method to handle transitions between text and image tokens during sequence generation. This approach is used in both dataset construction and sequence generation (Figure \ref{fig:approach_generation}).

For dataset construction, we use a 2D input sequence method with a 3x3 context window around each token. We use special tokens to mark transitions between text and images, and add padding to keep context sizes consistent. For text, only previous tokens are included in the context windows. The \texttt{input\_seq\_construct} function implements this process.

During sequence generation, the \texttt{\_process\_token} function manages transitions. It handles text tokens (bytes) and image tokens (tuples) differently, and takes care of special characters like newlines and tabs. A \texttt{sequence\_clock} keeps track of the generation progress. Each new token's context is represented in a 3x3 matrix, which helps maintain coherence when switching between text and images.

\section{Training and Evaluation}

\subsection{Model Architectures}
We evaluated three compact model architectures, each with fewer than 100,000 parameters: a Recurrent Neural Network (RNN) using bidirectional Long Short-Term Memory (LSTM) units \cite{schmidhuber1997long}, a Transformer \cite{vaswani2017attention}, and Mamba, based on the State Space Model (SSM) \cite{dao2024transformers}. These models were adapted to process our dataset of pixel data and bytecode sequences. Training was conducted on Kaggle using dual T4 GPUs, with a batch size of 32, sequence length of 256, stride size of 32, and learning rate of 0.001. We trained for 200 epochs, evaluating performance every 5 epochs. The resulting models are available at \url{https://huggingface.co/ffurfaro/PixelBytes-Pokemon}. Additionally, we developed a more specialized model for generation tasks, which is not presented in this paper.

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{training_results.png}
\caption{Training and validation metrics (Loss, Accuracy) for RNN, Transformer, and SSM models over 200 epochs.}
\label{fig:training_results}
\end{figure}

Figure \ref{fig:training_results} shows the training and validation metrics for our three model types over 200 epochs. The State Space Models (SSM) achieved the best scores for loss and accuracy. However, the widening gap between their training and validation curves suggests they may be overfitting, meaning they learned the training data well but struggled to generalize to new examples.

In contrast, the LSTMs (referred to as RNNs in our analysis) demonstrated more balanced performance. The closer alignment of their training and validation curves indicates they may generalize better to unseen data. The Transformer model had the lowest performance among the three. This could be due to the absence of positional encoding trick, which is important for Transformers in sequence tasks.

Overall, these results highlight the strengths and weaknesses of each model type. While SSMs showed strong training performance, their tendency to overfit suggests that additional regularization techniques might be needed. The stable performance of RNNs indicates their suitability for our task, as they balance well between fitting the training data and generalizing to new data. The Transformer's performance suggests it may require further adjustments to improve its effectiveness for this task.

\subsection{Generation Evaluation Metrics}

We tested different model types including State Space Models (SSM), Attention models (Att), and Recurrent Neural Networks (RNN) for generating 16 consecutive sequences. We used three metrics to evaluate their performance: Hamming Distance, Cosine Similarity, and BLEU Score \cite{sutskever2014sequence, hamming1950error, papineni2002bleu}.

\begin{table}[h]
\centering
\small
\begin{tabular}{cccccccccc}
\hline
Type & Dir. & Emb. & Conv. & In Emb & Hidden State & Depth & Hamming & Cosine & BLEU \\
\hline
SSM & Bi & PxBy & Y & 81 & 64 & 1 & 0.170 $\pm$ 0.086 & 0.883 $\pm$ 0.105 & 0.753 $\pm$ 0.115 \\
SSM & Bi & PxBy & Y & 81 & 64 & 2 & 0.158 $\pm$ 0.074 & 0.896 $\pm$ 0.095 & 0.771 $\pm$ 0.097 \\
SSM & Uni & PxBy & Y & 81 & 64 & 2 & 0.166 $\pm$ 0.081 & 0.886 $\pm$ 0.102 & 0.760 $\pm$ 0.106 \\
Att & - & PxBy & Y & 81 & 64 & 1 & 0.157 $\pm$ 0.064 & 0.887 $\pm$ 0.103 & 0.765 $\pm$ 0.088 \\
Att & - & PxBy & Y & 81 & 64 & 2 & 0.159 $\pm$ 0.066 & 0.887 $\pm$ 0.103 & 0.760 $\pm$ 0.092 \\
RNN & Bi & Center & N & 81 & 64 & 2 & 0.185 $\pm$ 0.074 & 0.888 $\pm$ 0.083 & 0.750 $\pm$ 0.093 \\
RNN & Bi & PxBy & Y & 81 & 64 & 2 & 0.153 $\pm$ 0.061 & 0.902 $\pm$ 0.090 & 0.777 $\pm$ 0.083 \\
RNN & Bi & PxBy & Y & 162 & 64 & 2 & 0.152 $\pm$ 0.062 & 0.905 $\pm$ 0.089 & 0.778 $\pm$ 0.084 \\
RNN & Bi & PxBy & Y & 36 & 64 & 2 & 0.152 $\pm$ 0.061 & 0.904 $\pm$ 0.090 & 0.778 $\pm$ 0.083 \\
RNN & Bi & PxBy & Y & 81 & 128 & 2 & 0.153 $\pm$ 0.063 & 0.903 $\pm$ 0.091 & 0.776 $\pm$ 0.086 \\
RNN & Bi & PxBy & Y & 81 & 32 & 2 & \textbf{0.149} $\pm$ 0.062 & 0.899 $\pm$ 0.095 & \textbf{0.785} $\pm$ 0.082 \\
RNN & Bi & PxBy & Y & 81 & 64 & 1 & 0.149 $\pm$ 0.062 & 0.897 $\pm$ 0.096 & 0.780 $\pm$ 0.085 \\
RNN & Bi & PxBy & Y & 81 & 64 & 3 & 0.153 $\pm$ 0.063 & \textbf{0.906} $\pm$ 0.087 & 0.776 $\pm$ 0.086 \\
RNN & Bi & PxBy & N & 81 & 64 & 2 & 0.151 $\pm$ 0.062 & 0.903 $\pm$ 0.090 & 0.779 $\pm$ 0.084 \\
RNN & Uni & PxBy & Y & 81 & 64 & 2 & 0.153 $\pm$ 0.064 & 0.904 $\pm$ 0.088 & 0.777 $\pm$ 0.087 \\
\hline
\end{tabular}
\caption{Comparison of model characteristics and performance (mean $\pm$ std)}
\label{tab:model_comparison}
\end{table}

The results show that RNN models with PxBy embedding generally perform better than other types. The best Hamming distance (0.149 $\pm$ 0.062) and BLEU score (0.785 $\pm$ 0.082) come from the bidirectional PxBy RNN model with convolution, 81 embedding dimensions, and 32 hidden state dimensions. This suggests that a smaller hidden state in the RNN model can effectively capture sequence patterns.

The highest cosine similarity (0.906 $\pm$ 0.087) is from the 3-layer RNN model, indicating that more layers can improve sequence alignment. Adding convolution usually improves RNN model performance. Interestingly, changing the embedding dimension (36, 81, 162) in RNN models didn't affect performance much, suggesting these models can handle different embedding sizes well. The center embedding in RNNs didn't perform as well as PxBy embedding for this task.

\section{Results and Discussion}

Our experiments with various model architectures tested the effectiveness of unified text and image generation. Bidirectional RNN models using PixelBytes (PxBy) embedding with convolutional layers generally performed better across our metrics. The PxBy embedding improved the mean BLEU score from 0.750 to 0.777 compared to a center-only approach, indicating its effectiveness in representing multimodal data. This finding aligns with recent work emphasizing the importance of effective multimodal embeddings in text-to-image generation tasks \cite{verHo2021efficient}.

Bidirectional models showed a slight advantage over unidirectional ones, with BLEU scores of 0.777 and 0.776 respectively. This suggests that bidirectionality may help capture sequential dependencies, consistent with other studies in sequence modeling \cite{schuster1997bidirectional}. Notably, varying state dimensions (32, 64, 128) had minimal impact on performance, with BLEU scores ranging from 0.775 to 0.784. While RNN models showed strong performance in our experiments, State Space Models (SSM) demonstrated rapid convergence during training, suggesting potential for future exploration in multimodal data processing. This is in line with recent advancements in SSM for sequence modeling tasks \cite{dao2024transformers}.

\section{Conclusion and Future Work}

This study on PixelBytes highlights the potential for unified text-image generation, contributing to the growing field of multimodal AI \cite{baltruvsaitis2018multimodal}. The performance of bidirectional RNN models with PxBy embedding underscores the importance of capturing both spatial and contextual information in multimodal data. Future work could focus on refining the PxByEmbed algorithm, testing on larger and more diverse datasets, exploring creative applications of multimodal generation, and further optimizing SSM and Attention models. These efforts aim to advance multimodal sequence modeling and generation techniques.

>>>>

\let\thefootnote\relax\footnotetext{$^{*}$ Corresponding author: Fabien Furfaro (\textbf{\href{mailto:fabien.furfaro@gmail.com}{\texttt{fabien.furfaro@capgemini.com}}}).}

\begin{abstract}
This report introduces PixelBytes, a novel approach for unified multimodal representation learning. Inspired by existing sequence models such as Image Transformers, PixelCNN, and Mamba-Bytes, our method aims to capture diverse inputs in a cohesive representation, exploring the integration of different data types, particularly text, audio, and pixelated images (sprites). We conducted our experiments on a specialized PixelBytes Pokémon dataset. Initially, we investigated various model architectures, including Recurrent Neural Networks (RNNs), State Space Models (SSMs), and Attention-based models, focusing on bidirectional processing and our convolutional PxBy embedding technique. Subsequently, we shifted our focus to evaluating models based on data reduction strategies and the effectiveness of autoregressive learning. We specifically examined Long Short-Term Memory (LSTM) networks in both predictive and autoregressive modes for our main experiments. Our findings indicate that autoregressive models outperform predictive models. By adopting a flexible approach to multimodal modeling, PixelBytes aims to support the development of foundation models capable of understanding and generating multimodal data. Code and pretrained models are available at \url{https://github.com/fabienfrfr/PixelBytes} and \url{https://huggingface.co/ffurfaro/aPixelBytes-Pokemon}.
\end{abstract}

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\linewidth]{aPxBy_approach.pdf}
\caption{Overview of the PixelBytes approach: (Left) The scanning process reads different modalities at specific positions. (Center) An example of our dataset and the autoregressive generation across modalities. (Right) A generation example, with the PxBy embedding window displayed below. Note: The model currently shows some inaccuracies in image generation and may produce incorrect words, highlighting areas for future improvement.}
\label{fig:approach_overview}
\end{figure}

@@

Recent advancements in artificial intelligence have led to increasingly generalist models, not by combining multiple specialized components (like Gato from DeepMind \cite{reed2022generalist}), but by giving simple tasks to models where emergent properties—complex behaviors that arise from simpler underlying rules—appear. This is the case with generative language models like GPT \cite{brown2020language}. However, these models are limited by their focus on language alone, failing to capture the full complexity of multimodal understanding \cite{huang2023language}. To address this limitation, researchers have explored combining LLMs with other modalities \cite{liu2024visual}. But this brings us back to the initial problem, as it often results in specialized model combinations without allowing for new emergent properties. We propose "PixelBytes", a novel approach enabling unified training across modalities by capturing diverse inputs in a single, cohesive representation.

Multimodal sequence generation, which involves the creation of coherent outputs combining various data types such as text, images, and numerical sequences, presents a significant challenge in artificial intelligence \cite{baltruvsaitis2018multimodal}. While models like GPT have excelled in text generation \cite{brown2020language}, there's a growing need for unified approaches that can seamlessly handle varied data types. Building upon these findings, PixelBytes aims to address the challenge of unified text, audio and image generation by proposing a model capable of producing mixed sequences of text, audio and images in a coherent and unified manner. We draw inspiration from state-of-the-art sequence models, including Image Transformers \cite{parmar2018image}, PixelCNN \cite{van2016conditional}, and recent developments in bytes generation by mamba architectures \cite{wang2024mambabyte}.

Our research explores various architectures including Recurrent Neural Networks (RNNs), State Space Models (SSMs) \cite{dao2024transformers}, and Attention-based models, focusing on the effectiveness of bidirectional processing \cite{zhu2024vision}, novel embedding techniques (particularly PxBy embedding, which unifies pixel and byte-level representations), the impact of convolutional layers, effects of model depth, input dimensionality, and size.

Our experiments, conducted on a specialized PixelBytes Pokémon dataset, reveal that autoregressive models with balanced data reduction strategies outperform predictive models in terms of accuracy and loss. While our initial exploration suggested potential for bidirectional sequence models with PxBy embedding and convolutional layers, our final results focus on the performance of Long Short-Term Memory (LSTM) networks in both predictive and autoregressive modes. This paper presents our methodology for constructing unified sequence data from text, audio, and pixelated images (sprites), as well as our experimental results and analysis. By adopting a flexible approach to multimodal modeling, PixelBytes aims to advance the development of versatile foundation models capable of understanding and generating multimodal data.

\section{Exploration for a Unified Representation}

\subsection{Hypothesis Testing Framework}

The quest for a unified data representation across different modalities presents significant challenges. Text data typically has a one-dimensional dependency on preceding words with discrete values \cite{bengio2003neural}. Audio signals, on the other hand, have a temporal dimension with continuous values \cite{van2016wavenet}. Action-state representations in robotics can be similar to audio but with correlations between channels \cite{botteghi2021low}. Images and animations combine spatial and temporal dimensions across discrete RGB channels \cite{krizhevsky2012imagenet}. Given these diverse characteristics, many researchers have focused on combining separate embedding models into a single framework, as seen in projects like ImageBind \cite{girdhar2023imagebind} and RT-2 \cite{zitkovich2023rt}, rather than seeking a truly unified data representation. However, to build a model that comprehensively understands different modalities without intermediate alignment steps, exploring a unified data representation becomes necessary. In this section, we will test several hypotheses:

\begin{itemize}
    \item Can we quantize data such that each element becomes a token? \cite{zhan2022auto}
    \item Is predicting only the next value sufficient for a sequence model to learn effectively?
    \item For dimensions higher than one, is applying a convolutional filter necessary?
    \item For space-time dependency, what is the importance of bidirectionality in models?
\end{itemize}

Through these investigations, we aim to explore a method of representing data that could enable models to understand various types of modalities.

\subsection{Conceptual Multimodal Embedding}

\subsubsection{Dataset Construction}

To test our hypotheses on unified representation, we needed a dataset combining visual and textual data in a format suitable for byte-level processing. Image captioning datasets proved unsuitable due to limited text content and difficulties in interpreting pixelated versions of high-resolution images. We therefore created a specialized Pokémon dataset, offering pixelated designs and rich descriptive text. We collected data by web scraping Pokémon miniatures and descriptions from Pokepedia using Beautiful Soup \cite{richardson2007beautiful}, maintaining a 2/3 text to 1/3 image ratio. For image processing, we employed a 55-color palette inspired by the NES, creating tokens for different color combinations. This approach allowed us to represent visual information in a format compatible with our tokenizer.

\sloppy

To handle transitions between text and image tokens, we developed a 2D input sequence method using a 3x3 context window around each token with a 2D zigzag scheme (Figure \ref{fig:approach_overview}). Special tokens mark transitions between text and images, with padding added to maintain consistent context sizes. The padding value is 0. For text, only previous tokens are included in the context windows. Using OpenCV and scikit-image \cite{bradski2000opencv, van2014scikit} for image quantization and pixelization, we adjusted all entries to have 113 indices, balancing text and image tokens. The resulting dataset, combining text and pixelated images, is available on the Hugging Face Datasets Hub (\url{https://huggingface.co/datasets/ffurfaro/PixelBytes-Pokemon}) for reproducibility. It includes a "pixelbyte" column for this specific data representation.

\subsubsection{Embedding Techniques}

Our exploration of unified representation begins with integrating image-text data for sequence generation. We developed a tokenizer that processes "pixelbyte" column. This is paired with an embedding technique we call PxByEmbed, which creates a unified representation for pixel and byte data in a single space. We designed PxByEmbed to test our hypotheses about the effectiveness of convolutional filters for higher-dimensional data. PxByEmbed uses a learned embedding matrix to map each token (text or image) to a vector space, while maintaining spatial relationships for image tokens. It incorporates a simple convolutional layer and an adaptive mixing mechanism. This design allows us to investigate if predicting only the next value (with or without only previous value) is sufficient for effective learning in a sequence model.

\begin{algorithm}[ht]
\footnotesize
\caption{PxByEmbed: Multimodal Embedding Algorithm (k=3)
\newline
\textbf{Input:} $V$: vocabulary size, $D$: embedding dimension
\newline
\textbf{Output:} Embedded representation $\mathbf{E} \in \mathbb{R}^{B \times L \times D}$
\newline
\textbf{Note:} $\mathbf{X}_{emb} \in \mathbb{R}^{B \cdot L \times E_{int} \times k \times k}$, 
$\mathbf{X}_{flat} \in \mathbb{R}^{B \cdot L \times E_{int}k^2}$, 
$\mathbf{X}_{proj} \in \mathbb{R}^{B \cdot L \times D}$
}
\begin{algorithmic}[0]
\State \textbf{Initialize:}
\State $k \gets 3$
\State $E_{int} \gets \max(9, \lfloor D / k^2 \rfloor)$
\State $\mathbf{\alpha} \in \mathbb{R}^{1 \times 1 \times k \times k}$ 
\State $\mathbf{W}_{emb} \in \mathbb{R}^{V \times E_{int}}$ 
\State $\mathbf{W}_{proj} \in \mathbb{R}^{E_{int}k^2 \times D}$ 
\State $\mathbf{W}_{patch} \in \mathbb{R}^{E_{int} \times E_{int} \times k \times k}$

@@

The algorithm processes input sequences in 3x3 patches, embedding each element and then applying a learnable convolution. The embedding pad value is set to 0 to avoid influencing the training of subsequent tokens. These two representations are then combined using a learned parameter, allowing the model to adaptively balance between local and global information.

\subsection{Model Architectures Evaluated}

We evaluated three compact model architectures: a Recurrent Neural Network (RNN) using Long Short-Term Memory (LSTM) units \cite{schmidhuber1997long}, a Transformer \cite{vaswani2017attention}, and a State Space Model (SSM) based on Mamba \cite{gu2023mamba}. For both the RNN and Mamba architectures, we compared versions with and without a bidirectional first layer. Each model was constrained to fewer than 100,000 parameters and adapted to process our dataset of pixel data and bytecode sequences. We trained the models on Kaggle using T4 GPUs, with a batch size of 32, sequence length of 256, and learning rate of 0.001 for 200 epochs. The resulting models are available at \url{https://huggingface.co/ffurfaro/PixelBytes-Pokemon}.

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{training_results_embed.pdf}
\caption{Training and validation metrics for RNN, Transformer, and SSM models over 200 epochs.}
\label{fig:training_results}
\end{figure}

\subsection{Comparative Analysis}

Figure \ref{fig:training_results} shows the training and validation metrics for our three model types. The SSM achieved the best scores for loss and accuracy but showed signs of overfitting. The RNN demonstrated more balanced performance, suggesting better generalization to unseen data. The Transformer had the lowest performance, possibly due to the absence of positional encoding. The strong performance of the SSM supports the potential of unified representation for pixel and byte data. However, its tendency to overfit suggests that predicting only the next value might not be sufficient for effective learning in all cases. The balanced performance of the RNN indicates that simpler architectures can still be effective for our task. The Transformer's lower performance suggests that complex attention mechanisms may not always be necessary or beneficial for this type of data.

\subsubsection{Generation Evaluation Metrics}

To assess the effectiveness of our approach and various model architectures, we evaluated their generation capabilities. We tested State Space Models (SSM), Attention models (Att), and Recurrent Neural Networks (RNN) in generating 32 consecutive sequences. Our evaluation used three metrics: Hamming Distance \cite{hamming1950error}, Cosine Similarity, and BLEU Score \cite{papineni2002bleu}.

\begin{table}[ht]
\centering
\small
\begin{tabular}{cccccccccc}
\hline
Type & Dir. & Emb. & Conv. & In Emb & Hidden State & Depth & Hamming & Cosine & BLEU \\
\hline
SSM & Bi & PxBy & Y & 81 & 64 & 1 & 0.170 $\pm$ 0.086 & 0.883 $\pm$ 0.105 & 0.753 $\pm$ 0.115 \\
SSM & Bi & PxBy & Y & 81 & 64 & 2 & 0.158 $\pm$ 0.074 & 0.896 $\pm$ 0.095 & 0.771 $\pm$ 0.097 \\
SSM & Uni & PxBy & Y & 81 & 64 & 2 & 0.166 $\pm$ 0.081 & 0.886 $\pm$ 0.102 & 0.760 $\pm$ 0.106 \\
Att & - & PxBy & Y & 81 & 64 & 1 & 0.157 $\pm$ 0.064 & 0.887 $\pm$ 0.103 & 0.765 $\pm$ 0.088 \\
Att & - & PxBy & Y & 81 & 64 & 2 & 0.159 $\pm$ 0.066 & 0.887 $\pm$ 0.103 & 0.760 $\pm$ 0.092 \\
RNN & Bi & Center & N & 81 & 64 & 2 & 0.185 $\pm$ 0.074 & 0.888 $\pm$ 0.083 & 0.750 $\pm$ 0.093 \\
RNN & Bi & PxBy & Y & 81 & 64 & 2 & 0.153 $\pm$ 0.061 & 0.902 $\pm$ 0.090 & 0.777 $\pm$ 0.083 \\
RNN & Bi & PxBy & Y & 162 & 64 & 2 & 0.152 $\pm$ 0.062 & 0.905 $\pm$ 0.089 & 0.778 $\pm$ 0.084 \\
RNN & Bi & PxBy & Y & 36 & 64 & 2 & 0.152 $\pm$ 0.061 & 0.904 $\pm$ 0.090 & 0.778 $\pm$ 0.083 \\
RNN & Bi & PxBy & Y & 81 & 128 & 2 & 0.153 $\pm$ 0.063 & 0.903 $\pm$ 0.091 & 0.776 $\pm$ 0.086 \\
RNN & Bi & PxBy & Y & 81 & 32 & 2 & \textbf{0.149} $\pm$ 0.062 & 0.899 $\pm$ 0.095 & \textbf{0.785} $\pm$ 0.082 \\
RNN & Bi & PxBy & Y & 81 & 64 & 1 & 0.149 $\pm$ 0.062 & 0.897 $\pm$ 0.096 & 0.780 $\pm$ 0.085 \\
RNN & Bi & PxBy & Y & 81 & 64 & 3 & 0.153 $\pm$ 0.063 & \textbf{0.906} $\pm$ 0.087 & 0.776 $\pm$ 0.086 \\
RNN & Bi & PxBy & N & 81 & 64 & 2 & 0.151 $\pm$ 0.062 & 0.903 $\pm$ 0.090 & 0.779 $\pm$ 0.084 \\
RNN & Uni & PxBy & Y & 81 & 64 & 2 & 0.153 $\pm$ 0.064 & 0.904 $\pm$ 0.088 & 0.777 $\pm$ 0.087 \\
\hline
\end{tabular}
\caption{Comparison of model characteristics and performance (mean $\pm$ std)}
\label{tab:model_comparison}
\end{table}

The results show variations across different model configurations. Among RNN models with PxBy embedding, we observed a Hamming distance of 0.149 $\pm$ 0.062 and a BLEU score of 0.785 $\pm$ 0.082 for a bidirectional model with convolution, 81 embedding dimensions, and 32 hidden state dimensions. The highest cosine similarity (0.906 $\pm$ 0.087) was achieved by a 3-layer RNN model. We noted some differences in performance between models with and without convolution, as well as between unidirectional and bidirectional configurations, though these differences were not always substantial. Changing the embedding dimension (36, 81, 162) in RNN models resulted in similar performance levels. RNN models using center embedding showed different results compared to those using PxBy embedding. The performance of SSM and Attention models varied in comparison to RNN models across the different metrics, but no model type consistently outperformed the others across all measures.

\subsection{Identified Challenges}

Our initial results revealed several limitations in our embedding approach. While we observed variations in performance across different model configurations, the differences were often not substantial \cite{wang2020deep}. The PxBy embedding, which we initially thought promising, did not consistently outperform simpler approaches across all metrics and model types. Based on these findings, we recognized the need to refine our approach. The repetition of sequences in our generated output indicated that our embedding method might not be capturing the full range of patterns in our data \cite{bengio2013representation}.

\section{Optimizing Unified Representation}

\subsection{Refined Embedding Approach}

To address the challenges identified in our initial experiments, we propose a revised embedding strategy. Instead of using a convolutional approach, we now focus on six specific positions within each token, with the input dimension equal to the output dimension. This change allows for a larger embedding size while potentially improving the model's ability to capture relevant patterns. We also recognized the need for a more flexible tokenizer \cite{kudo2018sentencepiece}. Our initial implementation became cumbersome when generating new data, highlighting the importance of a more versatile approach. We are exploring ways to integrate all necessary functionality into the tokenizer itself, which should simplify our overall pipeline and potentially improve performance. These refinements reflect a shift from our initial exploration towards a more focused approach to unified representation. While our initial results provided valuable insights, they also revealed the complexities inherent in multimodal sequence modeling and the need for continuous iteration in our methods \cite{lecun2015deep}.

\subsubsection{Dataset Construction}

\sloppy

For our refined approach, we created a new dataset combining images, text, and audio extracted from Pokemon sprite animations. This dataset, available at \url{https://huggingface.co/datasets/ffurfaro/PixelBytes-PokemonAll}, was compiled through web scraping of Pokepedia and includes descriptions of the Pokemon along with their associated cries. The dataset consists of animated, pixelated GIFs of Pokemon sprites as the visual component. The audio files are two-channel recordings: Channel 1 contains the original mono sound of the Pokemon cry, while Channel 2 features a filtered version simulating a bits Game Boy speaker output to verify our approach for control problems. This setup allows us to model a simplified dynamic physical system, where the original sound acts as the "action" input and the filtered output represents the "state" of the system. The transfer function of this bandpass filter can be approximated as:

\begin{equation}
H(s) = \frac{K \omega_n^2}{s^2 + 2\zeta\omega_n s + \omega_n^2}
\end{equation}

where $K$ is the gain, $\omega_n$ is the natural frequency, and $\zeta$ is the damping ratio. These parameters can be adjusted to closely match the frequency response of a Game Boy speaker.

\subsection{Enhanced Tokenization Strategy}

Building on our previous work, we developed an improved tokenization strategy using the ActionPixelBytesTokenizer. This tokenizer aims to handle multimodal data more effectively, including text, images, and audio, while maintaining a unified representation. The tokenizer employs a combined vocabulary that includes ASCII bytes, RGB values from the NES palette, and action states for control and audio. This approach attempts to create a consistent representation across different data types. For text processing, the tokenizer converts to lowercase ASCII bytes. Images are converted to Lab color space and quantized to the nearest NES palette color. Audio data is normalized and mapped to predefined action states, were setpoint is reset to zero (standard equilibrium). All token numbers are now 151, where index 0 corresponds to the null padding value, and indices 1 and 2 are transition values. A key aspect of the new tokenizer is its sequence construction method. Instead of using convolutional methods, we focus on six specific positions for each token for no repetition in sequencing of 3D zigzag scheme (Figure \ref{fig:approach_overview}). This approach creates context-target pairs that aim to capture relationships between neighboring tokens in both space and time. The sequence construction algorithm is detailed below:
